\bibitem{refDavletovEn}
Davletov, A. R. Modern methods of machine learning and OCR technology for automating document processing / A. R. Davletov. – Text : electronic // Bulletin of Science. – 2023. – No. 10 (67). – URL: https://cyberleninka.ru/article/n/sovremennye-metody-mashinnogo-obucheniya-i-tehnologiya-ocr-dlya-avtomatizatsii-obrabotki-dokumentov (accessed: 22.05.2026).

\bibitem{refKuchuganovEn}
Kuchuganov, A. V. Methodology of semantic analysis and search for graphic information : abstract of the dissertation for the degree of Doctor of Technical Sciences : specialty 05.13.01 "System analysis, management and information processing (in industry)" / A. V. Kuchuganov. – Izhevsk, 2018. – 32 p. – Text : direct.

\bibitem{refRakovEn}
Rakov, N. S. Machine vision / N. S. Rakov, S. V. Palmov. – Text : electronic // Forum of Young Scientists. – 2018. – No. 4 (20). – URL: https://cyberleninka.ru/article/n/mashinnoe-zrenie (accessed: 22.05.2026).

\bibitem{refRublevskyEn}
Rublevsky, R. The landscape of key terms in generative AI, their relationship and use / R. Rublevsky. – Text : electronic // Habr. – 2025. – 25 Sept. – URL: https://habr.com/ru/articles/950600/ (accessed: 22.05.2026).

\bibitem{refStarovoytovEn}
Starovoytov, V. V. Image acquisition and processing on computers : educational and methodological manual / V. V. Starovoytov, Yu. I. Golub. – Minsk : BNTU, 2018. – 204 p. – ISBN 978-985-550-770-4. – Text : direct.

\bibitem{refTrofimovEn}
Trofimov, I. L. The Role of Libraries in the Era of Artificial Intelligence: An Analytical Perspective on New AI Technologies and Their Applications / I. L. Trofimov, E. M. Kustova, S. M. Barash, A. V. Buryak, V. N. Filatova // Modern Technologies. System Analysis. Modeling. – 2026. – No. 3(87). – P. 94–104. – URL: https://ojs.irgups.ru/index.php/stsam/article/view/2450.

\bibitem{refAkhilEn}
Akhil, S. Overview of Tesseract OCR engine / S. Akhil. – Text : electronic // ResearchGate. – 2016. – Dec. – URL: https://www.researchgate.net/publication/315834331 (accessed: 22.05.2026).

\bibitem{refLineAngleEn}
Chapter 1 -- Organic Chemistry Review / Hydrocarbons. Section 1.4 Condensed Structural and Line-Angle Formulas. – Text : electronic // Hostos Community College Library Guides. – URL: https://guides.hostos.cuny.edu/che120 (accessed: 22.05.2026).

\bibitem{refDalbyEn}
Dalby, A. Description of Several Chemical Structure File Formats Used by Computer Programs Developed at Molecular Design Limited / A. Dalby, J. G. Nourse, W. D. Hounshell [et al.] // Journal of Chemical Information and Computer Sciences. – 1992. – Vol. 32, no. 3. – P. 244–255. – DOI: 10.1021/ci00007a012.

\bibitem{refCAMPEn}
Fifty, C. In-Context Learning for Few-Shot Molecular Property Prediction / C. Fifty, J. Leskovec, S. Thrun. – Text : electronic // arXiv.org. – 2023. – arXiv:2310.08863. – URL: https://arxiv.org/abs/2310.08863 (accessed: 22.05.2026).

\bibitem{refGoFEn}
Gamma, E. Design Patterns: Elements of Reusable Object-Oriented Software / E. Gamma, R. Helm, R. Johnson, J. Vlissides. – Reading : Addison-Wesley, 1995. – 395 p. – ISBN 0-201-63361-2. – Text : direct.

\bibitem{refXYCutEn}
Ha, J. Recursive X-Y Cut Using Bounding Boxes of Connected Components / J. Ha, R. M. Haralick, I. T. Phillips // Proceedings of the Third International Conference on Document Analysis and Recognition. – 1995. – Vol. 2. – P. 952–955. – DOI: 10.1109/ICDAR.1995.602058.

\bibitem{refInChIEn}
Heller, S. R. InChI, the IUPAC International Chemical Identifier / S. R. Heller, A. McNaught, I. Pletnev [et al.] // Journal of Cheminformatics. – 2015. – Vol. 7. – Art. 23. – DOI: 10.1186/s13321-015-0068-4.

\bibitem{refLlavaEn}
Liu, H. Visual Instruction Tuning / H. Liu, C. Li, Q. Wu, Y. J. Lee. – Text : electronic // arXiv.org. – 2023. – arXiv:2305.11845. – URL: https://arxiv.org/abs/2305.11845 (accessed: 22.05.2026).

\bibitem{refOBoyleCanonicalEn}
O'Boyle, N. M. Towards a Universal SMILES Representation – A Standard Method to Generate Canonical SMILES Based on the InChI / N. M. O'Boyle // Journal of Cheminformatics. – 2012. – Vol. 4. – Art. 22. – DOI: 10.1186/1758-2946-4-22.

\bibitem{refUMLEn}
OMG Unified Modeling Language (OMG UML), Version 2.5.1 / Object Management Group. – Text : electronic. – 2017. – URL: https://www.omg.org/spec/UML/2.5.1/PDF (accessed: 22.05.2026).

\bibitem{refWikiOCSREn}
Optical chemical structure recognition. – Text : electronic // Wikipedia. – URL: https://en.wikipedia.org/wiki/Optical\_chemical\_structure\_recognition (accessed: 22.05.2026).

\bibitem{refMolScribeEn}
Qian, Y. MolScribe: Robust Molecular Structure Recognition with Image-To-Graph Generation / Y. Qian, J. Guo, Z. Tu [et al.]. – Text : electronic // arXiv.org. – 2022. – arXiv:2205.14311. – URL: https://arxiv.org/pdf/2205.14311 (accessed: 22.05.2026).

\bibitem{refQwen3VLEn}
Qwen Team. Qwen3-VL Technical Report / Qwen Team. – Text : electronic // arXiv.org. – 2025. – arXiv:2511.21631. – URL: https://arxiv.org/abs/2511.21631 (accessed: 22.05.2026).

\bibitem{refDECIMEREn}
Rajan, K. DECIMER 1.0: deep learning for chemical image recognition using transformers / K. Rajan [et al.]. – Text : electronic // ResearchGate. – 2021. – Aug. (accessed: 22.05.2026).

\bibitem{refDECIMERJournalEn}
Rajan, K. DECIMER 1.0: deep learning for chemical image recognition using transformers / K. Rajan, A. Zielesny, C. Steinbeck. – Text : electronic // Journal of Cheminformatics. – 2021. – Vol. 13. – DOI: 10.1186/s13321-021-00538-8. – URL: https://doi.org/10.1186/s13321-021-00538-8 (accessed: 22.05.2026).

\bibitem{refLayoutParserEn}
Shen, Z. LayoutParser: A Unified Toolkit for Deep Learning Based Document Image Analysis / Z. Shen [et al.]. – Text : electronic // arXiv.org. – 2021. – arXiv:2103.15348. – URL: https://arxiv.org/abs/2103.15348 (accessed: 22.05.2026).

\bibitem{refOCRMetricsEn}
Smith, R. An Overview of the Tesseract OCR Engine / R. Smith // Ninth International Conference on Document Analysis and Recognition (ICDAR 2007). – Washington : IEEE Computer Society, 2007. – Vol. 2. – P. 629–633. – DOI: 10.1109/ICDAR.2007.4376991. – URL: https://doi.org/10.1109/ICDAR.2007.4376991 (accessed: 22.05.2026).

\bibitem{refSmithEn}
Smith, R. An Overview of the Tesseract OCR Engine / R. Smith // Ninth International Conference on Document Analysis and Recognition (ICDAR 2007). – Washington : IEEE Computer Society, 2007. – Vol. 2. – P. 629–633. – DOI: 10.1109/ICDAR.2007.4376991. – URL: https://doi.org/10.1109/ICDAR.2007.4376991 (accessed: 22.05.2026).

\bibitem{refTrinajsticEn}
Trinajstić, N. Chemical Graph Theory / N. Trinajstić. – 2nd ed. – Boca Raton : CRC Press, 1992. – 364 p. – Text : direct.

\bibitem{refPEP8En}
van Rossum, G. PEP 8 -- Style Guide for Python Code / G. van Rossum, B. Warsaw, A. Coghlan. – Text : electronic // Python.org. – URL: https://peps.python.org/pep-0008/ (accessed: 23.05.2026).

\bibitem{refMinerUEn}
Wang, B. MinerU: An Open-Source Solution for Precise Document Content Extraction / B. Wang, C. Xu, X. Zhao, L. Ouyang, F. Wu, Z. Zhao, R. Xu, K. Liu, Y. Qu, F. Shang [et al.] // arXiv preprint. – 2024. – arXiv:2409.18839. – DOI: 10.48550/arXiv.2409.18839.

\bibitem{refWarrEn}
Warr, W. A. Representation of Chemical Structures / W. A. Warr // Wiley Interdisciplinary Reviews: Computational Molecular Science. – 2011. – Vol. 1, no. 4. – P. 557–579. – DOI: 10.1002/wcms.36.

\bibitem{refWeininger1En}
Weininger, D. SMILES, a Chemical Language and Information System. 1. Introduction to Methodology and Encoding Rules / D. Weininger // Journal of Chemical Information and Computer Sciences. – 1988. – Vol. 28, no. 1. – P. 31–36. – DOI: 10.1021/ci00057a005.

\bibitem{refWeininger2En}
Weininger, D. SMILES. 2. Algorithm for Generation of Unique SMILES Notation / D. Weininger, A. Weininger, J. L. Weininger // Journal of Chemical Information and Computer Sciences. – 1989. – Vol. 29, no. 2. – P. 97–101. – DOI: 10.1021/ci00062a008.

\bibitem{refRLSAEn}
Wong, K. Y. Document Analysis System / K. Y. Wong, R. G. Casey, F. M. Wahl // IBM Journal of Research and Development. – 1982. – Vol. 26, no. 6. – P. 647–656. – DOI: 10.1147/rd.266.0647.

\bibitem{refLayoutLMEn}
Xu, Y. LayoutLM: Pre-training of Text and Layout for Document Image Understanding / Y. Xu, M. Li, L. Cui [et al.]. – Text : electronic // arXiv.org. – 2019. – arXiv:1912.13318. – URL: https://arxiv.org/abs/1912.13318 (accessed: 22.05.2026).
